---
title: "Parks"
author: "Callan Hoskins"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 150)

# Libraries
library(tidyverse)
library(viridis)
library(lubridate)
library(dcl)
library(ggenealogy)

# Parameters
acquistions_file_path <- here::here("data/acquisitions.rds")
all_cetacean_data_file_path <- here::here("data/all_cetaceans_data.rds")
na_names <- 
  c(
    "(hybrid)", 
    "(not named)", 
    "(no name)", 
    "x", 
    "(none)", 
    "#", 
    "@ riverhead\rfound."
  )

# this function standardizes some records that signify the same thing but are
# represented by different names/strings. Part of data cleaning
standardize_locs <- function(location) {
  location <- 
    location %>% 
    str_replace_all("[\\(\\)\\*\\.]|(\\? ?)", "")
  case_when(
    str_detect(location, "Dolphin Connection") | 
      str_detect(location, "Dinnes") ~ "Dolphin Connection", 
    str_detect(location, "Dolphin Quest Hawai") ~ "Dolphin Quest Hawaii", 
    str_detect(location, "(Dolphin Research Center|DRC)") ~ 
      "Dolphin Research Center", 
    str_detect(location, "Discovery Cove") ~ "Discovery Cove",
    str_detect(location, "EPCOT") ~ "EPCOT Center", 
    str_detect(location, "Gulf World Marine Park") ~ "Gulf World Marine Park", 
    str_detect(location, "Gulfarium") ~ "Gulfarium Marine Adventure Park", 
    str_detect(location, "(Marine Animal Productions|MAP)") ~ 
      "Marine Animal Productions", 
    str_detect(location, "Inst.*Marine Mammal Studies") ~ 
      "Institute of Marine Mammal Studies", 
    str_detect(location, "Audubon") ~ "Audubon Zoo", 
    str_detect(location, "Kahala.*Oriental.*") ~ 
      "Kahala Mandarin Oriental Hawaii",
    str_detect(location, "Kamogawa SeaWorld") ~ "SeaWorld Kamogawa", 
    str_detect(location, "^Marine Life") ~ "Marine Life Oceanarium", 
    str_detect(location, "Minnesota Zoo") ~ "Minnesota Zoo", 
    str_detect(location, "National Aquarium") ~ "National Aquarium Baltimore", 
    str_detect(location, "Niag.*") ~ "Niagara Aquarium", 
    str_detect(location, "Point Defiance Zoo") ~ "Point Defiance Zoo", 
    str_detect(location, "Sea Life Park") ~ "Sea Life Park Hawaii", 
    str_detect(location, "Sea-Arama") ~ "Sea-Arama Marineworld", 
    str_detect(location, "Sealand.*Cape.*Cod.*") ~ "Sealand Cape Cod", 
    str_detect(location, "Seattle.*Aquarium") ~ "Seattle Marine Aquarium", 
    str_detect(location, "SeaWorld Aur*") ~ "SeaWorld Aurora", 
    str_detect(location, "Shedd") ~ "Shedd Aquarium", 
    str_detect(location, "SeaWorld.*Diego.*") ~ "SeaWorld San Diego", 
    str_detect(location, "Texas.*Stranding Network") ~ "TMMSN", 
    str_detect(location, "Texas State.*Aquarium.*") ~ "Texas State Aquarium", 
    str_detect(location, "Dolphinaris") ~ "Dolphinaris Arizona", 
    str_detect(location, "Navy, U\\.?S\\.?") ~ "US Navy", 
    str_detect(location, "Great America") ~ "Six Flags Great America", 
    str_detect(location, "^La Feria") ~ "La Feria de Chapultepec", 
    str_detect(location, "Six Flags Marine World") ~ "Marine World Africa USA", 
    str_detect(location, "Utrish") ~ "Utrish Aquarium, Moscow", 
    TRUE ~ location
  )
}

# Natural origin/destination locations
natural_locs <- 
  c(
    "Argentina", 
    "Atlantic Ocean", 
    "Atlantic Ocean, Unknown", 
    "Avalon Beach, Florida", 
    "^Bolivar", 
    "Bullfrog Creek, Florida", 
    "California, unknown beach", 
    "CaliforniaPacific Ocean", 
    "Carr Inlet", 
    "Cedar Point", 
    "Chile, Straits of Magellan", 
    "Churchill Oaks, Louisiana", 
    "Clearwater Beach, Florida", 
    "Compano Bay, Texas", 
    "Corpus Christi, Texas", 
    "Cudjoe Key, Florida", 
    "Curacao, Caribbean", 
    "Eastern Florida", 
    "Floating Pen, Grindavík Harbor", 
    "Florida", 
    "Florida Keys", 
    "Escape, Virgin Islands", 
    "Florida Keys, Escape", 
    "Florida Keys, Re-Capture", 
    "Florida, Atlantic Ocean", 
    "Florida, Atlantic Ocean, off New Smyrna", 
    "Florida, Gulf Coast", 
    "Florida, Gulf of Mexico", 
    "Florida, location unknown", 
    "Florida, unknown", 
    "Florida, Unknown", 
    "Fort Pierce", 
    "Fourchon", 
    "Gulf of California", 
    "Gulf of Mexico", 
    "Gulf Shores", 
    "Hatakejiriwan", 
    "Hawaii, Pacific Ocean", 
    "[hH]olding", 
    "Hudson Bay, Manitoba", 
    "Iceland", 
    "Indian River Lagoon, Florida", 
    "Indian River", 
    "Jacksonville, Florida", 
    "Japan,", 
    "^Kettenburg", 
    "Lower Matecumbe Key", 
    "Melbourne Beach, Florida", 
    "Mississippi, Gulf of Mexico", 
    "Moriya coast, Japan", 
    "Mosquito Lagoon, Florida", 
    "Mustang Island", 
    "Near Key West", 
    "Okhotsk Sea, Russia", 
    "[Ee]scape", 
    "Pacific Ocean, California", 
    "Panama City, Florida", 
    "Pedder Bay", 
    "Pender Harbor", 
    "Penn Cove", 
    "^Ponce", 
    "^Port", 
    "^Private", 
    "^Puget Sound", 
    "^Quintana Beach", 
    "[cC]apture", 
    "^Redington Beach", 
    "^Sanibel", 
    "Sea [pP]en", 
    "^Shell Island", 
    "South Padre Island", 
    "^Straits of Magellan", 
    "^Tampa Bay", 
    "^Texas, Gulf", 
    "^Three Sisters", 
    "^[Uu]nknown", 
    "Ultrish", 
    "Vero Beach, Florida",
    "White Sea, Russia"
  )

```

# Families within Captive Cetacean Populations

```{r}
# Read in data on individual cetaceans
all_cetacean_data <- 
  read_rds(all_cetacean_data_file_path)
```

We now want to inspect some of the family ties between cetaceans in captivity. But first, we have to make sure that our data is clean. 

## Data cleaning

These animals are connected not by unique identifiers but by their names, some of which may be spelled incorrectly. We have to make sure that each `name` is unique to one animal.

```{r}
all_cetacean_data <-
  all_cetacean_data %>% 
  rename(
    "id" = X1
  ) %>% 
  mutate(
    name = str_to_lower(name), 
    mother = str_to_lower(mother), 
    father = str_to_lower(father)
  ) %>% 
  filter(!name %in% na_names)
```

```{r}
duplicate_entries <-
  all_cetacean_data %>% 
  filter(!str_detect(name, "no |fetus|none|unk")) %>% 
  mutate(
    name = case_when(
      name == "(joe)" ~ "joe", 
      name == "(rosalie)" ~ "rosalie", 
      TRUE ~ name
    )
  ) %>% 
  count(name, species, entry_date, status_date) %>% 
  filter(n >= 2) %>% 
  pull(name)

duplicate_entries
```

There are almost 20 animals that have more than one entry for their `name` / `entry_date` combination. I will visually inspect these duplicates and select the entry for each that seems most accurate. 

```{r}
all_cetacean_data %>% 
  filter(name %in% duplicate_entries) %>% 
  select(name, id, entry_date, official_id, everything()) %>% 
  arrange(name, entry_date)
```

I will have to fix some of this data manually. I had to find some updates online. Here is a summary of my changes: 

* "Alia" the bottlenose dolphin passed away at Dolphinaris Arizona on May 22, 2018. I am combining her information into one updated entry. 
* "Bodie" the bottlenose dolphin passed away at Dolphinaris Arizona on September 23, 2017. I am combining his information into one updated entry. 
* There is a second bottlenose dolphin named "Bodie" who was born in 2014 and is currently at Seaworld San Diego who should still be included in the data. 
* "Kai'nalu" the bottlenose dolphin passed away at Dolphinaris Arizona on January 31, 2019. I am combining his information into one updated entry. 
* "Khloe" the bottlenose dolphin passed away at Dolphinaris Arizona on December 31, 2019. I am combining her information into one updated entry. 
* Neither of the two entries for "Lazarus" seems to give more information. I am arbitrarily dropping the entry with `id = 1810`. 
* "Liko" the bottlenose dolphin at Dolphinaris Arizona is currently still alive and at Dolphinaris Arizona. I will update this information, collected 2/2/2019. I will drop Liko's unknown `status` entry. 
* There is a second bottlenose dolphin named "Liko" unrelated to the one at Dolphinaris Arizona. I will keep his entry as-is. 
* The entry for "Mirabella" with `id = 1691` has information about Mirabella's mother and father, so I will drop the other one (`id = 1069`). 
* There are two nearly-identical entries for a dolphin named "Misty" who was born 1980-07-27 and died 2005-01-18. One of the entries says that his father was "Spray" and the other one has `NA` for that entry, so we will choose the more complete entry (remove `id = 2011`). 
* "Noelani" the dolphin at Dolphinaris Arizona was reportedly alive 2/2/2019. I will combine her repetitive entries into one updated one. 
* There is a second, unrelated "Noelani" who I will not edit. 
* "Orion" the bottlenose dolphin born in 1982 seems to have more information in the entry with `id = 780`. I will drop his entry with `id = 1656`. 
* There is a second, seemingly unrelated bottlenose dolphin named "Orion" whose information I will not edit. 
* "Ping" of Dolphinaris Arizona was reportedly alive 2/2/2019. I will combine her repetitive entries into one. 
* "Sonny" the bottlenose dolphin of Dolphinaris Arizona was reportedly alive 2/2/2019. I will combine his repetitive entries into one updated entry and leave the rest of the separate "Sonny" entries alone. 
* There are three repetitive entries for the bottlenose dolphin named "Spanky," and one is far more informative than the others. I will remove the two that are uninformative. 
* There are two repetitive entries for the bottlenose dolphin named "Splash," and one is more informative than the other. I will remove the one that is uninformative. The remaining dolphins named "Splash" are unique. 
* There are two repetitive entries for the bottlenose dolphin named "Spray," and one is more informative than the other. I will remove the one that is uninformative. The remaining dolphins named "Spray" are unique. 

```{r}
all_cetacean_data <-
  all_cetacean_data %>% 
  filter(
    !(name == "alia" & status == "Unknown"), # drop one of alia's entries
    !(name == "bodie" & id != 304 & status == "Unknown"), # drop one of bodie1's entries
    !(name == "kai'nalu" & id != 603), # drop one of kai'nalu's entries
    !(name == "khloe" & id != 608), # drop one of khloe's entries, 
    !id == 1810, # drop one of Lazarus's entries, 
    !(name == "liko" & status == "Unknown"), # drop one of liko's entries
    !(name == "mirabella" & id == 1069), # drop one of Mirabella's entries
    !(name == "misty" & id == 2011), # drop one of misty's entries
    !(name == "noelani" & status == "Unknown"), # drop one of noelani's entries
    !(name == "orion" & id == 1656), # drop one of orion's entries
    !(name == "ping" & status == "Unknown"), # drop one of ping's entries
    !(name == "sonny" & status == "Unkown" & id == 605), # drop one of sonny's entries
    !(name == "spanky" & id != 751), # remove two repetitive entries for spanky
    !(name == "splash" & id == 1807), # remove a repetitive entry for splash
    !(name == "spray" & id == 2176)
  ) %>% 
  mutate(
    status = if_else(name == "alia", "Died", status), # update alia's info
    status_date = 
      if_else(name == "alia", make_date(2018L, 5L, 22L), status_date), # update alia's info
    status = if_else(name == "bodie", "Died", status), # update bodie's info
    status_date = 
      if_else(name == "bodie", make_date(2017L, 9L, 23L), status_date), # update bodie's info
    status = if_else(name == "kai'nalu", "Died", status), # update kai'nalu's info
    status_date = 
      if_else(name == "kai'nalu", make_date(2019L, 1L, 31L), status_date), #update kai'nalu's info
    status = if_else(name == "khloe", "Died", status), # update khloe's info
    status_date = 
      if_else(name == "khloe", make_date(2018L, 12L, 1L), status_date), #update khloe's info
    status = if_else(name == "liko", "Alive", status), # update liko's status
    status_date = 
      if_else(name == "liko", make_date(2019L, 2L, 2L), status_date), # update liko's status
    status_date = 
      if_else(
        name == "noelani" & id == 241, 
        make_date(2019L, 2L, 2L), 
        status_date
      ), # update noelani's info
    status = if_else(name == "ping", "Alive", status), # update ping's info
    status_date = 
      if_else(name == "ping", make_date(2019L, 2L, 2L), status_date), # update ping's info
    status_date = 
      if_else(
        name == "sonny" & id == 213, 
        make_date(2019L, 2L, 2L), 
        status_date
      ), # update sonny's info
  ) %>% 
  distinct(name, .keep_all = TRUE)
```

In case you were wondering about the number of deaths/unknown information at Dolphinaris Arizona, the facility closed in 2019 due to the fact that four of their eight dolphins died within the first two years of their opening. You can read about it [here](!https://www.abc15.com/news/region-northeast-valley/scottsdale/dolphinaris-arizona-temporarily-closing-after-death-of-facilitys-fourth-dolphin). 

# Parks

Another of the most potentially interesting things about this data is the opportunity to inspect the track records of many parks that keep cetaceans. As popular films like *Blackfish* and *Free Willy* have exposed, the conditions that cetaceans live in in parks like Seaworld is not good enough for them to thrive. We saw an example of this mismanagement (or at least bad luck) in the data I corrected regarding the dolphins from Dolphinaris Arizona, which lost four of its eight dolphins within four years of opening. What parks are the best / worst with regards to how their animals live and breed? I am not a marine biologist, but I know that lifespan and breeding record are two rudimentary ways to measure an animal's quality of life in captivity. 

```{r}
transfer_data <- 
  all_cetacean_data %>% 
  select(
    -c(
      transfer, 
      transfer_date, 
      notes, region, 
      status_date, 
      cod, 
      mother, 
      father, 
      accuracy, 
      id, 
      official_id
    )
  ) %>% 
  # standardize spellings of Hawaii
  mutate_at(
    vars(origin_loc, currently, transfers), 
    str_replace_all, 
    "Hawai'i", 
    "Hawaii"
  ) %>% 
  # standardize spellings of SeaWorld
  mutate_at(
    vars(origin_loc, currently, transfers), 
    str_replace_all, 
    "[Ss]ea[ ]?[Ww]orld", 
    "SeaWorld"
  )

transfer_data %>% 
  pull(transfers) %>% 
  head()
```

The first problem is how to extract information about these transfers from these strings... we are going to need some serious regex here!

```{r}
transfer_data <- 
  transfer_data %>% 
  mutate(location = str_split(transfers, " to ")) %>% 
  unnest(location) %>% 
  mutate(
    location = 
      case_when(
        is.na(location) & origin_loc == currently ~ origin_loc, 
        is.na(location) & origin_loc == "Unknown" ~ currently, 
        name == "hapa" ~ "U.S. Navy MMP",
        TRUE ~ location
      ), 
    transfer_year = 
      if_else(
        str_detect(location, "\\d{4}"), 
        str_extract(location, "\\d{4}") %>% as.integer(), 
        origin_date %>% year() %>% as.integer()
      ), 
    location = str_replace(location, " \\(.*\\).*", "")
  ) %>% 
  mutate_at(vars(location, origin_loc, currently), standardize_locs) %>% 
  filter(location != "") %>% 
  group_by(name) %>% 
  mutate(transfer_no = seq(1, n())) %>% 
  ungroup() %>% 
  select(
    transfer_no, 
    name, 
    location, 
    transfers, 
    origin_loc, 
    currently, 
    everything()
  )
```

## Cetaceans with high numbers of transfers

```{r}
transfer_data %>% 
  group_by(name) %>% 
  summarize(n_locs = n()) %>% 
  ggplot(aes(n_locs)) + 
  geom_histogram(binwidth = 1)
```

We can see that the vast majority of captive cetaceans were transfered less than five times in their lives. Let's take a look at the cetaceans that were transfered the most: 

```{r}
transfer_data %>% 
  group_by(name) %>% 
  mutate(n_locs = n()) %>% 
  filter(n_locs > 10) %>% 
  select(name, species, origin_date, birth_year, entry_date, status, acquisition, n_locs) %>% 
  distinct()
```

This data certainly tells a sad story. Many of these animals lived only a few years before dying, and they were transfered so many times within that short span of time. Orcas Kahana, Kilroy, and Kenau lived a year or less in captivity before passing away, during which time they were transferred more than ten times. Again, I am no marine biologist, but I recognize that a transfer puts incredible stress on an animal. 

## Cetacean Lifetime by Location

I am interested in seeing how long animals live when they are housed in certain parks. Our naive first approach below seems to have the fault that it does not make a difference between the collection site and the ultimate place where the animals end up. This doesn't provide much information about the parks because we have no knowledge or control over the place where the animals were originally captured. 

```{r}
transfer_data %>% 
  mutate(capt_lifetime = (birth_year %--% entry_date) %>% time_length("years")) %>% 
  drop_na(capt_lifetime) %>% 
  group_by(location) %>% 
  summarize(med_capt_lifetime = median(capt_lifetime)) %>% 
  top_n(20, med_capt_lifetime) %>% 
  ggplot(aes(fct_reorder(location, med_capt_lifetime), med_capt_lifetime)) + 
  geom_col() + 
  coord_flip()
```

```{r}
transfer_data_parks <- 
  transfer_data %>% 
  filter(!map_lgl(location, ~ str_detect(., natural_locs) %>% any()))
transfer_data_parks %>% 
  count(location) %>% 
  arrange(location)
```

```{r}
transfer_data_parks %>% 
  mutate(capt_lifetime = (birth_year %--% entry_date) %>% time_length("years")) %>% 
  drop_na(capt_lifetime) %>% 
  group_by(location) %>% 
  summarize(med_capt_lifetime = median(capt_lifetime)) %>% 
  top_n(20, med_capt_lifetime) %>% 
  ggplot(aes(fct_reorder(location, med_capt_lifetime), med_capt_lifetime)) + 
  geom_point() + 
  coord_flip() + 
  scale_y_continuous(breaks = scales::breaks_width(5)) + 
  theme_light() + 
  labs(
    title = "Twenty Best Parks for Median Cetacean Lifetime", 
    y = "Median Cetacean Lifetime (years)", 
    x = NULL
  )
```

