---
title: "Progress Report 2"
author: "Callan Hoskins"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 150)

# Libraries
library(tidyverse)
library(viridis)
library(lubridate)
library(dcl)

# Parameters
acquistions_file_path <- here::here("data/acquisitions.rds")
all_cetacean_data_file_path <- here::here("data/all_cetaceans_data.rds")
```

# Families within Captive Cetacean Populations

```{r}
# Read in data on individual cetaceans
all_cetacean_data <- 
  read_rds(all_cetacean_data_file_path)
```

We now want to inspect some of the family ties between cetaceans in captivity. But first, we have to make sure that our data is clean. 

## Data cleaning

These animals are connected not by unique identifiers but by their names, some of which may be spelled incorrectly. We have to make sure that each `name` is unique to one animal.

```{r}
all_cetacean_data <-
  all_cetacean_data %>% 
  rename(
    "id" = X1
  ) %>% 
  mutate(
    name = str_to_lower(name), 
    mother = str_to_lower(mother), 
    father = str_to_lower(father)
  ) 
```

```{r}
duplicate_entries <-
  all_cetacean_data %>% 
  filter(
    !name %in% c("(hybrid)", "(not named)", "(no name)", "x", "(none)", "#", "@ riverhead\rfound."), 
    !str_detect(name, "no |fetus|none|unk")
  ) %>% 
  mutate(
    name = case_when(
      name == "(joe)" ~ "joe", 
      name == "(rosalie)" ~ "rosalie", 
      TRUE ~ name
    )
  ) %>% 
  count(name, species, entry_date, status_date) %>% 
  filter(n >= 2) %>% 
  pull(name)

duplicate_entries
```

There are almost 20 animals that have more than one entry for their `name` / `entry_date` combination. I will visually inspect these duplicates and select the entry for each that seems most accurate. 

```{r}
all_cetacean_data %>% 
  filter(name %in% duplicate_entries) %>% 
  select(name, id, entry_date, official_id, everything()) %>% 
  arrange(name, entry_date)
```

I will have to fix some of this data manually. I had to find some updates online. Here is a summary of my changes: 

* "Alia" the bottlenose dolphin passed away at Dolphinaris Arizona on May 22, 2018. I am combining her information into one updated entry. 
* "Bodie" the bottlenose dolphin passed away at Dolphinaris Arizona on September 23, 2017. I am combining his information into one updated entry. 
* There is a second bottlenose dolphin named "Bodie" who was born in 2014 and is currently at Seaworld San Diego who should still be included in the data. 
* "Kai'nalu" the bottlenose dolphin passed away at Dolphinaris Arizona on January 31, 2019. I am combining his information into one updated entry. 
* "Khloe" the bottlenose dolphin passed away at Dolphinaris Arizona on December 31, 2019. I am combining her information into one updated entry. 
* Neither of the two entries for "Lazarus" seems to give more information. I am arbitrarily dropping the entry with `id = 1810`. 
* "Liko" the bottlenose dolphin at Dolphinaris Arizona is currently still alive and at Dolphinaris Arizona. I will update this information, collected 2/2/2019. I will drop Liko's unknown `status` entry. 
* There is a second bottlenose dolphin named "Liko" unrelated to the one at Dolphinaris Arizona. I will keep his entry as-is. 
* The entry for "Mirabella" with `id = 1691` has information about Mirabella's mother and father, so I will drop the other one (`id = 1069`). 
* There are two nearly-identical entries for a dolphin named "Misty" who was born 1980-07-27 and died 2005-01-18. One of the entries says that his father was "Spray" and the other one has `NA` for that entry, so we will choose the more complete entry (remove `id = 2011`). 
* "Noelani" the dolphin at Dolphinaris Arizona was reportedly alive 2/2/2019. I will combine her repetitive entries into one updated one. 
* There is a second, unrelated "Noelani" who I will not edit. 
* "Orion" the bottlenose dolphin born in 1982 seems to have more information in the entry with `id = 780`. I will drop his entry with `id = 1656`. 
* There is a second, seemingly unrelated bottlenose dolphin named "Orion" whose information I will not edit. 
* "Ping" of Dolphinaris Arizona was reportedly alive 2/2/2019. I will combine her repetitive entries into one. 
* "Sonny" the bottlenose dolphin of Dolphinaris Arizona was reportedly alive 2/2/2019. I will combine his repetitive entries into one updated entry and leave the rest of the separate "Sonny" entries alone. 
* There are three repetitive entries for the bottlenose dolphin named "Spanky," and one is far more informative than the others. I will remove the two that are uninformative. 
* There are two repetitive entries for the bottlenose dolphin named "Splash," and one is more informative than the other. I will remove the one that is uninformative. The remaining dolphins named "Splash" are unique. 
* There are two repetitive entries for the bottlenose dolphin named "Spray," and one is more informative than the other. I will remove the one that is uninformative. The remaining dolphins named "Spray" are unique. 

```{r}
all_cetacean_data <-
  all_cetacean_data %>% 
  filter(
    !(name == "alia" & status == "Unknown"), # drop one of alia's entries
    !(name == "bodie" & id != 304 & status == "Unknown"), # drop one of bodie1's entries
    !(name == "kai'nalu" & id != 603), # drop one of kai'nalu's entries
    !(name == "khloe" & id != 608), # drop one of khloe's entries, 
    !id == 1810, # drop one of Lazarus's entries, 
    !(name == "liko" & status == "Unknown"), # drop one of liko's entries
    !(name == "mirabella" & id == 1069), # drop one of Mirabella's entries
    !(name == "misty" & id == 2011), # drop one of misty's entries
    !(name == "noelani" & status == "Unknown"), # drop one of noelani's entries
    !(name == "orion" & id == 1656), # drop one of orion's entries
    !(name == "ping" & status == "Unknown"), # drop one of ping's entries
    !(name == "sonny" & status == "Unkown" & id == 605), # drop one of sonny's entries
    !(name == "spanky" & id != 751), # remove two repetitive entries for spanky
    !(name == "splash" & id == 1807), # remove a repetitive entry for splash
    !(name == "spray" & id == 2176)
  ) %>% 
  mutate(
    status = if_else(name == "alia", "Died", status), # update alia's info
    status_date = 
      if_else(name == "alia", make_date(2018L, 5L, 22L), status_date), # update alia's info
    status = if_else(name == "bodie", "Died", status), # update bodie's info
    status_date = 
      if_else(name == "bodie", make_date(2017L, 9L, 23L), status_date), # update bodie's info
    status = if_else(name == "kai'nalu", "Died", status), # update kai'nalu's info
    status_date = 
      if_else(name == "kai'nalu", make_date(2019L, 1L, 31L), status_date), #update kai'nalu's info
    status = if_else(name == "khloe", "Died", status), # update khloe's info
    status_date = 
      if_else(name == "khloe", make_date(2018L, 12L, 1L), status_date), #update khloe's info
    status = if_else(name == "liko", "Alive", status), # update liko's status
    status_date = 
      if_else(name == "liko", make_date(2019L, 2L, 2L), status_date), # update liko's status
    status_date = 
      if_else(
        name == "noelani" & id == 241, 
        make_date(2019L, 2L, 2L), 
        status_date
      ), # update noelani's info
    status = if_else(name == "ping", "Alive", status), # update ping's info
    status_date = 
      if_else(name == "ping", make_date(2019L, 2L, 2L), status_date), # update ping's info
    status_date = 
      if_else(
        name == "sonny" & id == 213, 
        make_date(2019L, 2L, 2L), 
        status_date
      ), # update sonny's info
  )
```

In case you were wondering about the number of deaths/unknown information at Dolphinaris Arizona, the facility closed in 2019 due to the fact that four of their eight dolphins died within the first two years of their opening. You can read about it [here](!https://www.abc15.com/news/region-northeast-valley/scottsdale/dolphinaris-arizona-temporarily-closing-after-death-of-facilitys-fourth-dolphin). 


## Finding each animal's number of progeny

As an exploratory analysis, we want to see if we can find the number of progeny born from each animal in the dataset. Let's write a function we can map to each cetacean that will give us the number of progeny that animal produced. 
```{r}
calc_num_progeny <- function(name_param, sex_param) {
  if (sex_param == "F") {
    all_cetacean_data %>% 
      filter(mother == name_param) %>% 
      nrow()
  } else {
    all_cetacean_data %>% 
      filter(father == name_param) %>% 
      nrow()
  }
}

fertile_cetaceans <-
  all_cetacean_data %>%
  transmute(
    name,
    species,
    num_progeny = map2_int(name, sex, calc_num_progeny), 
    sex,
    origin_date,
    acquisition
  ) %>% 
  filter(num_progeny > 0)

fertile_cetaceans
```

**Note**: The methodology for finding cetacean offspring is *flawed* and relies only on the names. To make this more robust, we have to factor in birthyear, transfer history, or some other metric to ensure that we don't overcount animals with more popular names. 

Let's try to visualize the captive cetaceans' breeding history. 

First, we want to get an idea of the distribution of each type species of cetacean in the dataset as a whole. 

```{r}
fertile_cetaceans %>% 
  ggplot(aes(fct_rev(fct_infreq(species)))) + 
  geom_bar() + 
  coord_flip()
```

We can see that a vast majority of the animals in the dataset are bottlenose dolphins -- so much of a majority, in fact, that they dwarf the population size of all the other species. Let's just focus on bottlenose dolphins, belugas, and killer whales/orcas for now. 

```{r}
common_species <- c("Killer Whale; Orca", "Bottlenose", "Beluga")

fertile_cetaceans %>% 
  filter(species %in% common_species) %>% 
  drop_na(origin_date, num_progeny) %>% 
  ggplot(aes(num_progeny)) + 
  geom_bar(aes(fill = species), position = "dodge", binwidth = 1) + 
  labs(
    title = "Number of Progeny by Species", 
    y = "Frequency", 
    x = "Number of progeny"
  )

fertile_cetaceans %>% 
  filter(species %in% common_species) %>% 
  drop_na(origin_date, num_progeny) %>% 
  ggplot(aes(num_progeny)) + 
  geom_bar(aes(fill = sex), position = "dodge", binwidth = 1) + 
  labs(
    title = "Number of progeny by sex", 
    y = "Frequency", 
    x = "Number of Progeny"
  )
```

From the first bar plot above, we can see that the distribution of number of progeny among the three most common species has a similar shape. From the second plot, we can see that, on the whole, female cetaceans tend bear more children than the fathers sire. However, one must note that the cetaceans with the greatest number of progeny fostered are all male, probably because of the ease with which caretakers can distribute male sperm to females. 

```{r}
fertile_cetaceans %>% 
  ggplot(aes(sex)) + 
  geom_bar()
```

We see from the plot above that there are actually far more female captive cetaceans than male ones, more than double! When we compare the number of progeny coming from each of the two sexes, we should probably be taking the overall number of cetaceans of that sex into account. 

```{r}
all_cetacean_data %>% 
  filter(
    !acquisition %in% c("Stillbirth", "Miscarriage", "Unknown"), 
    sex != "U"
  ) %>% 
  ggplot(aes(acquisition)) + 
  geom_bar(aes(fill = sex), position = "dodge")
```

The above graph shows us that the total number of cetaceans born and rescued is almost equally distributed between the two sexes, with slightly over a hundred more female cetaceans captured than male cetaceans. This does not seem to explain the incongruency between the number of cetaceans produced by male versus female cetaceans. 

## Inspecting Parks

Another of the most potentially interesting things about this data is the opportunity to inspect the track records of many parks that keep cetaceans. As popular films like *Blackfish* and *Free Willy* have exposed, the conditions that cetaceans live in in parks like Seaworld is not good enough for them to thrive. We saw an example of this mismanagement (or at least bad luck) in the data I corrected regarding the dolphins from Dolphinaris Arizona, which lost four of its eight dolphins within four years of opening. What parks are the best / worst with regards to how their animals live and breed? I am not a marine biologist, but I know that lifespan and breeding record are two rudimentary ways to measure an animal's quality of life in captivity. 
